---
title: "Meta-analysis physiological costs of hibernation: Report 3 - GC"
author: "Pablo Capilla-Lasheras"
date: "2025-01-29"
output:
  html_document:
    df_print: paged
editor_options: 
  chunk_output_type: console
knit: (function(inputFile, encoding) {
      out_dir <- "../05_reports";
      rmarkdown::render(inputFile,
                        encoding=encoding,
                        output_dir=file.path(dirname(inputFile), out_dir))})
---

```{r setup, include = FALSE}

# setup
knitr::opts_chunk$set(
  include = TRUE,
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)


library(openxlsx)
library(metafor)
library(tidyverse)
library(ggokabeito)
library(orchaRd)
library(rmarkdown)
library(knitr)
library(kableExtra)
library(gtsummary)
```

This [R Markdown](http://rmarkdown.rstudio.com) Notebook contains the initial meta-analysis for the project 'physiological 'costs' of hibernation.

## Dataset and calculation of effect sizes

The dataset for this meta-analysis has been collected after a literature search by Pablo Burraco, Marko Prokic and Tamara Petrovic. It contains data on physiological effects of hibernation in different systems. Here a quick snapshot of the first lines of the dataset:

```{r}
# dataset with raw data
data  <- openxlsx::read.xlsx("C:/Users/pca/Dropbox/PabloCL/2-Work drive/3-Active_projects/8 - Collaborations/Meta-analysis_hibernation/01_data/Hibernation_dataset_4_1_.xlsx",
                             colNames=T,
                             sheet = 3)

paged_table(data, options = list(rows.print = 20))
```

As a summary of the *initial dataset*:

-   Number of studies: `r length(unique(data$Study_ID))`
-   Number of species: `r length(unique(data$Species))`

For this meta-analysis, I am going to include all glucocorticoid (GC) measures and calculate the standarised mean difference (g) as the effect size of analysis comparing **hibernation** *versus* **euthermia**. Positive SMD values indicate higher GC values in *hibernation* while negative SMD indicate lower GC values in hibernation. This is a summary of the effect size calculation based on the dataset above:

```{r}
# ES data
data <- readRDS("C:/Users/pca/Dropbox/PabloCL/2-Work drive/3-Active_projects/8 - Collaborations/Meta-analysis_hibernation/01_data/processed_RDS_data_files/03_GC_metaanalysis_full_data.RDS")
data$obsID <- 1:nrow(data)

## data for SMD
df_SMD <- data %>%
  filter(!is.na(SMD)) %>% 
  filter(!is.na(SMD.sv))
```

Summary of SMD effect size and number of observations included in the meta-analysis:

```{r}
summary(df_SMD$SMD)


df_SMD %>% 
  filter(SMD > 5)

```

-   Number of observation with non-missing SMD: `r nrow(df_SMD)`
-   Number of studies in the dataset for analysis: `r length(unique(df_SMD$Study_ID))`
-   Number of species in the dataset for analysis: `r length(unique(df_SMD$Species))`

## Initial meta-analysis:

This initial model contains only the intercept and five random effects to account for variation between between classes, families, species, between studies and within studies beyond sampling variance.

```{r echo = TRUE, eval = TRUE}
model_SMD <- rma.mv(yi = SMD, 
                     V = SMD.sv, 
                     mods = ~ 1, 
                     random = list(~1|Study_ID,
                                   ~1|Class,
                                   ~1|Family,
                                   ~1|Species,
                                   ~1|obsID),
                     data=df_SMD, 
                     method="REML")
```

```{r}
saveRDS(object = model_SMD, file = 'C:/Users/pca/Dropbox/PabloCL/2-Work drive/3-Active_projects/8 - Collaborations/Meta-analysis_hibernation/04_models/01_hibernation/01c_metamodel_hibernation_GC.RDS')
model_SMD <- readRDS(file = 'C:/Users/pca/Dropbox/PabloCL/2-Work drive/3-Active_projects/8 - Collaborations/Meta-analysis_hibernation/04_models/01_hibernation/01c_metamodel_hibernation_GC.RDS')
summary(model_SMD)
i2_SMD <- round(i2_ml(model_SMD), digits = 4) # heterogeneity random effects
```

Proportion of variation left after accounting for sampling variance and proportion accounted by each random effect:

```{r}
knitr::kable(x = data.frame(Variable = names(i2_SMD),
                            i2_value_percentage = as.numeric(round(i2_SMD, digits = 3)))) %>% 
  kable_styling(full_width = F)
```

### Plot of results of intercept-only model:

```{r}
overall_plot <- orchard_plot(object = model_SMD, 
                             mod = 1, 
                             group = 'Study_ID',
                             trunk.size = 0.75,
                             alpha = 0.2,
                             cb = FALSE,
                             xlab = "Standardised Mean Difference (g)",
                             transfm = "none") +
  scale_fill_okabe_ito() +
  scale_color_okabe_ito() +
  theme(axis.text = element_text(size = 12),
        axis.title = element_text(size = 16))


ggsave(filename = "C:/Users/pca/Dropbox/PabloCL/2-Work drive/3-Active_projects/8 - Collaborations/Meta-analysis_hibernation/03_plots/01_hibernation/01c_overall_plot_hibernation_GC.png", 
       plot = overall_plot, 
       device = "png", 
       height = 90, 
       width = 150, 
       units = "mm")

overall_plot
```

## Meta-analysis including thermoregulation type as a moderator:

Number of observations for each type:

```{r}
table(df_SMD$Thermoregulation)
```

```{r, eval = TRUE}
model_SMD_thermoregulation <- rma.mv(yi = SMD, 
                                      V = SMD.sv, 
                                      mods = ~
                                        Thermoregulation - 
                                       1,
                                      random = list(~1|Study_ID,
                                                    ~1|Class,
                                                    ~1|Family,
                                                    ~1|Species,
                                                    ~1|obsID),
                                      data=df_SMD, 
                                      method="REML")


summary(model_SMD_thermoregulation)

```

```{r}
saveRDS(object = model_SMD_thermoregulation, file = 'C:/Users/pca/Dropbox/PabloCL/2-Work drive/3-Active_projects/8 - Collaborations/Meta-analysis_hibernation/04_models/01_hibernation/02c_metareg_thermoregulation_hibernation_GC.RDS')
model_SMD_thermoregulation <- readRDS(file = 'C:/Users/pca/Dropbox/PabloCL/2-Work drive/3-Active_projects/8 - Collaborations/Meta-analysis_hibernation/04_models/01_hibernation/02c_metareg_thermoregulation_hibernation_GC.RDS')
summary(model_SMD_thermoregulation)
```

### Plot of results of thermoregulation meta-regression:

```{r}
thermo_plot <- orchard_plot(object = model_SMD_thermoregulation, 
                            mod = 'Thermoregulation', 
                            group = 'Study_ID',
                            trunk.size = 0.75,
                            alpha = 0.2,
                            cb = FALSE,
                             xlab = "Standardised Mean Difference (g)",
                            transfm = "none") +
  scale_fill_okabe_ito() +
  scale_color_okabe_ito() +
  theme(axis.text = element_text(size = 12),
        axis.title = element_text(size = 16))


ggsave(filename = "C:/Users/pca/Dropbox/PabloCL/2-Work drive/3-Active_projects/8 - Collaborations/Meta-analysis_hibernation/03_plots/01_hibernation/02c_thermoregulation_plot_GC.png", 
       plot = thermo_plot, 
       device = "png", 
       height = 90, 
       width = 150, 
       units = "mm")

thermo_plot
```

