---
title: "Meta-analysis physiological costs of hibernation: Report 1 - OS"
author: "Pablo Capilla-Lasheras"
date: "2025-04-07"
output:
  html_document:
    df_print: paged
editor_options: 
  chunk_output_type: console
knit: (function(inputFile, encoding) {
      out_dir <- "../05_reports";
      rmarkdown::render(inputFile,
                        encoding=encoding,
                        output_dir=file.path(dirname(inputFile), out_dir))})
---

```{r setup, include = FALSE}

# setup
knitr::opts_chunk$set(
  include = TRUE,
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)


library(openxlsx)
library(metafor)
library(tidyverse)
library(ggokabeito)
library(orchaRd)
library(rmarkdown)
library(knitr)
library(kableExtra)
library(gtsummary)
```

This [R Markdown](http://rmarkdown.rstudio.com) Notebook contains the initial meta-analysis for the project 'physiological 'costs' of hibernation.

## Dataset and calculation of effect sizes

The dataset for this meta-analysis has been collected after a literature search by Pablo Burraco, Marko Prokic and Tamara Petrovic. It contains data on physiological effects of hibernation in different systems. Here a quick snapshot of the first lines of the dataset:

```{r}
# dataset with raw data
data  <- openxlsx::read.xlsx("C:/Users/pca/Dropbox/PabloCL/2-Work drive/3-Active_projects/8 - Collaborations/Meta-analysis_hibernation/meta_analysis_hibernation/01_data/Hibernation_dataset_4_1_.xlsx",
                             colNames=T,
                             sheet = 1)

paged_table(data, options = list(rows.print = 20))
```

As a summary of the *initial dataset*:

-   Number of studies: `r length(unique(data$Study_ID))`
-   Number of species: `r length(unique(data$Species))`

For this initial meta-analysis, I am going to include all oxidative stress (OS) markers and calculate the standarised mean difference (g; with heteroscedastic population variances) as the effect size of analysis comparing **hibernation** *versus* **euthermia**. Positive SMD values indicate higher OS values in *hibernation* while negative SMD indicate lower OS values in hibernation. This is a summary of the effect size calculation based on the dataset above:

```{r}
# ES data
data <- readRDS("C:/Users/pca/Dropbox/PabloCL/2-Work drive/3-Active_projects/8 - Collaborations/Meta-analysis_hibernation/meta_analysis_hibernation/01_data/processed_RDS_data_files/01_OS_metaanalysis_full_data.RDS")
data$obsID <- 1:nrow(data)

## data for SMD
df_SMD <- data %>%
  filter(!is.na(SMDH)) %>% 
  filter(!is.na(SMDH.sv))
```

Summary of SMD effect size and number of observations included in the meta-analysis:

```{r}
summary(df_SMD$SMDH)
```

-   Number of observation with non-missing SMD: `r nrow(df_SMD)`
-   Number of studies in the dataset for analysis: `r length(unique(df_SMD$Study_ID))`
-   Number of species in the dataset for analysis: `r length(unique(df_SMD$Species))`

## Initial meta-analysis:

This initial model contains only the intercept and five random effects to account for variation between between classes, families, species, between studies and within studies beyond sampling variance.

```{r echo = TRUE, eval = TRUE}
model_SMD <- rma.mv(yi = SMDH, 
                     V = SMDH.sv, 
                     mods = ~ 1, 
                     random = list(~1|Study_ID,
                                   ~1|Class,
                                   ~1|Family,
                                   ~1|Species,
                                   ~1|obsID),
                     data=df_SMD, 
                     method="REML")
```

```{r}
saveRDS(object = model_SMD, file = 'C:/Users/pca/Dropbox/PabloCL/2-Work drive/3-Active_projects/8 - Collaborations/Meta-analysis_hibernation/meta_analysis_hibernation/04_models/01_hibernation/01a_metamodel_hibernation_OS.RDS')
model_SMD <- readRDS(file = 'C:/Users/pca/Dropbox/PabloCL/2-Work drive/3-Active_projects/8 - Collaborations/Meta-analysis_hibernation/meta_analysis_hibernation/04_models/01_hibernation/01a_metamodel_hibernation_OS.RDS')
summary(model_SMD)
i2_SMD <- round(i2_ml(model_SMD), digits = 4) # heterogeneity random effects
```

Proportion of variation left after accounting for sampling variance and proportion accounted by each random effect:

```{r}
knitr::kable(x = data.frame(Variable = names(i2_SMD),
                            i2_value_percentage = as.numeric(round(i2_SMD, digits = 3)))) %>% 
  kable_styling(full_width = F)
```

### Plot of results of intercept-only model:

```{r}
overall_plot <- orchard_plot(object = model_SMD, 
                             mod = 1, 
                             group = 'Study_ID',
                             trunk.size = 1,
                             alpha = 0.2,
                             cb = FALSE,
                             xlab = "Standardised Mean Difference",
                             transfm = "none") +
  scale_fill_okabe_ito() +
  scale_color_okabe_ito() +
  theme(axis.text = element_text(size = 12),
        axis.text.y = element_text(size = 12, angle = 0), 
        axis.title = element_text(size = 16)) +
  scale_y_continuous(limits = c(-10, 15), breaks = seq(-10,15,5), label = seq(-10,15,5))


ggsave(filename = "C:/Users/pca/Dropbox/PabloCL/2-Work drive/3-Active_projects/8 - Collaborations/Meta-analysis_hibernation/meta_analysis_hibernation/03_plots/01_hibernation/01a_overall_plot_hibernation_OS.png", 
       plot = overall_plot, 
       device = "png", 
       height = 110, 
       width = 185, 
       units = "mm")

overall_plot
```

### Publication bias: small-study effects

```{r}
df_SMD$inv_eff_ss <- (1/df_SMD$Hib_N) + (1/df_SMD$Euthermia_N)


model_SMD_pb <- rma.mv(yi = SMDH, 
                     V = SMDH.sv, 
                     mods = ~ 1 + sqrt(inv_eff_ss), 
                     random = list(~1|Study_ID,
                                   ~1|Class,
                                   ~1|Family,
                                   ~1|Species,
                                   ~1|obsID),
                     data=df_SMD, 
                     method="REML")

summary(model_SMD_pb)
r2_ml(model_SMD_pb)

saveRDS(object = model_SMD_pb, file = 'C:/Users/pca/Dropbox/PabloCL/2-Work drive/3-Active_projects/8 - Collaborations/Meta-analysis_hibernation/meta_analysis_hibernation/04_models/01_hibernation/06a_pb_metamodel_hibernation_OS.RDS')
model_SMD_pb <- readRDS(file = 'C:/Users/pca/Dropbox/PabloCL/2-Work drive/3-Active_projects/8 - Collaborations/Meta-analysis_hibernation/meta_analysis_hibernation/04_models/01_hibernation/06a_pb_metamodel_hibernation_OS.RDS')
```

### Publication bias: time-lag effects

```{r}
model_SMD_tl <- rma.mv(yi = SMDH, 
                     V = SMDH.sv, 
                     mods = ~ 1 + scale(Year), 
                     random = list(~1|Study_ID,
                                   ~1|Class,
                                   ~1|Family,
                                   ~1|Species,
                                   ~1|obsID),
                     data=df_SMD, 
                     method="REML")

summary(model_SMD_tl)
r2_ml(model_SMD_tl)

saveRDS(object = model_SMD_tl, file = 'C:/Users/pca/Dropbox/PabloCL/2-Work drive/3-Active_projects/8 - Collaborations/Meta-analysis_hibernation/meta_analysis_hibernation/04_models/01_hibernation/07a_timelag_metamodel_hibernation_OS.RDS')
model_SMD_tl <- readRDS(file = 'C:/Users/pca/Dropbox/PabloCL/2-Work drive/3-Active_projects/8 - Collaborations/Meta-analysis_hibernation/meta_analysis_hibernation/04_models/01_hibernation/07a_timelag_metamodel_hibernation_OS.RDS')
```

## Meta-analysis with all organisms but including different OS markers:

Number of observations for each type:

```{r}
table(df_SMD$Biomarker_cat_2)
```

```{r, eval = TRUE}
model_SMD_biomarker <- rma.mv(yi = SMDH, 
                                      V = SMDH.sv, 
                                      mods = ~
                                        Biomarker_cat_2 - 
                                       1,
                                      random = list(~1|Study_ID,
                                                    ~1|Class,
                                                    ~1|Family,
                                                    ~1|Species,
                                                    ~1|obsID),
                                      data=df_SMD, 
                                      method="REML")


summary(model_SMD_biomarker)

```

```{r}
saveRDS(object = model_SMD_biomarker, file = 'C:/Users/pca/Dropbox/PabloCL/2-Work drive/3-Active_projects/8 - Collaborations/Meta-analysis_hibernation/meta_analysis_hibernation/04_models/01_hibernation/02a_metareg_biomarker_hibernation_OS.RDS')
model_SMD_biomarker <- readRDS(file = 'C:/Users/pca/Dropbox/PabloCL/2-Work drive/3-Active_projects/8 - Collaborations/Meta-analysis_hibernation/meta_analysis_hibernation/04_models/01_hibernation/02a_metareg_biomarker_hibernation_OS.RDS')
summary(model_SMD_biomarker)
```

### Plot of results of biomarker meta-regression:

```{r}
biomarker_plot <- orchard_plot(object = model_SMD_biomarker, 
                            mod = 'Biomarker_cat_2', 
                            group = 'Study_ID',
                            trunk.size = 1,
                            alpha = 0.2,
                            cb = FALSE,
                             xlab = "Standardised Mean Difference",
                            transfm = "none") +
  scale_fill_okabe_ito() +
  scale_color_okabe_ito() +
  theme(axis.text = element_text(size = 12),
        axis.text.y = element_text(size = 12, angle = 0), 
        axis.title = element_text(size = 16)) +
  scale_y_continuous(limits = c(-10, 18), breaks = seq(-10,15,5), label = seq(-10,15,5))


ggsave(filename = "C:/Users/pca/Dropbox/PabloCL/2-Work drive/3-Active_projects/8 - Collaborations/Meta-analysis_hibernation/meta_analysis_hibernation/03_plots/01_hibernation/02a_biomarker_OS.png", 
       plot = biomarker_plot, 
       device = "png", 
       height = 110, 
       width = 185, 
       units = "mm")

biomarker_plot
```


## Meta-analysis including thermoregulation type as a moderator:

Number of observations for each type:

```{r}
table(df_SMD$Thermoregulation)
```

```{r, eval = TRUE}
model_SMD_thermoregulation <- rma.mv(yi = SMDH, 
                                      V = SMDH.sv, 
                                      mods = ~
                                        Thermoregulation - 
                                       1,
                                      random = list(~1|Study_ID,
                                                    ~1|Class,
                                                    ~1|Family,
                                                    ~1|Species,
                                                    ~1|obsID),
                                      data=df_SMD, 
                                      method="REML")


summary(model_SMD_thermoregulation)

```

```{r}
saveRDS(object = model_SMD_thermoregulation, file = 'C:/Users/pca/Dropbox/PabloCL/2-Work drive/3-Active_projects/8 - Collaborations/Meta-analysis_hibernation/meta_analysis_hibernation/04_models/01_hibernation/03a_metareg_thermoregulation_hibernation_OS.RDS')
model_SMD_thermoregulation <- readRDS(file = 'C:/Users/pca/Dropbox/PabloCL/2-Work drive/3-Active_projects/8 - Collaborations/Meta-analysis_hibernation/meta_analysis_hibernation/04_models/01_hibernation/03a_metareg_thermoregulation_hibernation_OS.RDS')
summary(model_SMD_thermoregulation)
```

### Plot of results of thermoregulation meta-regression:

```{r}
thermo_plot <- orchard_plot(object = model_SMD_thermoregulation, 
                            mod = 'Thermoregulation', 
                            group = 'Study_ID',
                            trunk.size = 1,
                            alpha = 0.2,
                            cb = FALSE,
                             xlab = "Standardised Mean Difference",
                            transfm = "none") +
  scale_fill_okabe_ito() +
  scale_color_okabe_ito() +
  theme(axis.text = element_text(size = 12),
        axis.text.y = element_text(size = 12, angle = 0), 
        axis.title = element_text(size = 16)) +
    scale_y_continuous(limits = c(-10, 18), breaks = seq(-10,15,5), label = seq(-10,15,5))



ggsave(filename = "C:/Users/pca/Dropbox/PabloCL/2-Work drive/3-Active_projects/8 - Collaborations/Meta-analysis_hibernation/meta_analysis_hibernation/03_plots/01_hibernation/03a_thermoregulation_plot_OS.png", 
       plot = thermo_plot, 
       device = "png", 
       height = 120, 
       width = 185, 
       units = "mm")

thermo_plot
```

## Endotherms vs. ectotherms:

Now, we investigate if the response to hibernation varies across biomarkers of oxidative stress for endotherms and ectotherms. We include four markers:

```{r}
table(df_SMD$Biomarker_cat_2)
```

### Meta-model for endotherms:

```{r}
## data for SMD
df_SMD_endo <- data %>%
  filter(!is.na(SMDH)) %>% 
  filter(!is.na(SMDH.sv)) %>% 
  filter(Thermoregulation == 'Endothermic')

```

-   Number of observation with non-missing SMD: `r nrow(df_SMD_endo)`
-   Number of studies in the dataset for analysis: `r length(unique(df_SMD_endo$Study_ID))`
-   Number of species in the dataset for analysis: `r length(unique(df_SMD_endo$Species))`

```{r, eval = TRUE}
model_SMD_endo <- rma.mv(yi = SMDH, 
                          V = SMDH.sv, 
                          mods = ~
                            Biomarker_cat_2 -
                            1, 
                          random = list(~1|Study_ID,
                                        ~1|Family,
                                        ~1|Species,
                                        ~1|obsID),
                          data=df_SMD_endo, 
                          method="REML")


summary(model_SMD_endo)
```

```{r}
saveRDS(object = model_SMD_endo, file = 'C:/Users/pca/Dropbox/PabloCL/2-Work drive/3-Active_projects/8 - Collaborations/Meta-analysis_hibernation/meta_analysis_hibernation/04_models/01_hibernation/04a_metareg_hibernation_endo_OS.RDS')
model_SMD_endo <- readRDS(file = 'C:/Users/pca/Dropbox/PabloCL/2-Work drive/3-Active_projects/8 - Collaborations/Meta-analysis_hibernation/meta_analysis_hibernation/04_models/01_hibernation/04a_metareg_hibernation_endo_OS.RDS')
summary(model_SMD_endo)
```

```{r}
endo_plot <- orchard_plot(object = model_SMD_endo, 
                          mod = 'Biomarker_cat_2', 
                          group = 'Study_ID',
                          trunk.size = 1,
                          alpha = 0.2,
                          cb = FALSE,
                             xlab = "Standardised Mean Difference",
                          transfm = "none") +
  scale_fill_okabe_ito() +
  scale_color_okabe_ito() +
  theme(axis.text = element_text(size = 12),
        axis.text.y = element_text(size = 12, angle = 0), 
        axis.title = element_text(size = 16)) +
      scale_y_continuous(limits = c(-10, 18), breaks = seq(-10,15,5), label = seq(-10,15,5))



ggsave(filename = "C:/Users/pca/Dropbox/PabloCL/2-Work drive/3-Active_projects/8 - Collaborations/Meta-analysis_hibernation/meta_analysis_hibernation/03_plots/01_hibernation/04a_endo_plot_OS.png", 
       plot = endo_plot, 
       device = "png", 
       height = 120, 
       width = 185, 
       units = "mm")

endo_plot
```

### Meta-model for ectotherms:

```{r}
## data for SMD
df_SMD_ecto <- data %>%
  filter(!is.na(SMDH)) %>% 
  filter(!is.na(SMDH.sv)) %>% 
  filter(Thermoregulation == 'Ectothermic')

```

-   Number of observation with non-missing SMD: `r nrow(df_SMD_ecto)`
-   Number of studies in the dataset for analysis: `r length(unique(df_SMD_ecto$Study_ID))`
-   Number of species in the dataset for analysis: `r length(unique(df_SMD_ecto$Species))`

```{r, eval = TRUE}
model_SMD_ecto <- rma.mv(yi = SMDH, 
                          V = SMDH.sv, 
                          mods = ~
                            Biomarker_cat_2 -
                            1, 
                          random = list(~1|Study_ID,
                                        ~1|Class,
                                        ~1|Family,
                                        ~1|Species,
                                        ~1|obsID),
                          data=df_SMD_ecto, 
                          method="REML")
summary(model_SMD_ecto)
```

```{r}
saveRDS(object = model_SMD_ecto, file = 'C:/Users/pca/Dropbox/PabloCL/2-Work drive/3-Active_projects/8 - Collaborations/Meta-analysis_hibernation/meta_analysis_hibernation/04_models/01_hibernation/05a_metareg_ecto_hibernation_OS.RDS')
model_SMD_ecto <- readRDS(file = 'C:/Users/pca/Dropbox/PabloCL/2-Work drive/3-Active_projects/8 - Collaborations/Meta-analysis_hibernation/meta_analysis_hibernation/04_models/01_hibernation/05a_metareg_ecto_hibernation_OS.RDS')
summary(model_SMD_ecto)
```

```{r}
ecto_plot <- orchard_plot(object = model_SMD_ecto, 
                          mod = 'Biomarker_cat_2', 
                          group = 'Study_ID',
                          trunk.size = 0.75,
                          alpha = 0.2,
                          cb = FALSE,
                             xlab = "Standardised Mean Difference",
                          transfm = "none") +
  scale_fill_okabe_ito() +
  scale_color_okabe_ito() +
  theme(axis.text = element_text(size = 12),
        axis.text.y = element_text(size = 12, angle = 0), 
        axis.title = element_text(size = 16)) +
      scale_y_continuous(limits = c(-10, 18), breaks = seq(-10,15,5), label = seq(-10,15,5))

ggsave(filename = "C:/Users/pca/Dropbox/PabloCL/2-Work drive/3-Active_projects/8 - Collaborations/Meta-analysis_hibernation/meta_analysis_hibernation/03_plots/01_hibernation/05a_ecto_plot_OS.png", 
       plot = ecto_plot, 
       device = "png", 
       height = 120, 
       width = 185, 
       units = "mm")

ecto_plot
```
