---
title: "Meta-analysis physiological costs of hibernation: Report 6 - GC Arousal"
author: "Pablo Capilla-Lasheras"
date: "2026-05-04"
output:
  html_document:
    df_print: paged
editor_options: 
  chunk_output_type: console
knit: (function(inputFile, encoding) {
      out_dir <- "../05_reports";
      rmarkdown::render(inputFile,
                        encoding=encoding,
                        output_dir=file.path(dirname(inputFile), out_dir))})
---

```{r setup, include = FALSE}
# setup
knitr::opts_chunk$set(
  include = TRUE,
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)

library(here)
library(orchaRd)
library(openxlsx)
library(metafor)
library(tidyverse)
library(ggokabeito)
library(orchaRd)
library(rmarkdown)
library(knitr)
library(kableExtra)
library(gtsummary)
library(ggtree)
```

This [R Markdown](http://rmarkdown.rstudio.com) Notebook contains the initial meta-analysis for the project 'physiological 'costs' of hibernation. Specifically, the meta-analysis of arousal.

## Dataset and calculation of effect sizes

The dataset for this meta-analysis has been collected from the literature and curated by Pablo Burraco, Marko Prokic and Tamara Petrovic. It contains data on physiological effects of hibernation in different systems. Here a quick snapshot of the first lines of the dataset:

```{r echo = TRUE, eval = TRUE, include = TRUE}
# dataset with raw data
data <- openxlsx::read.xlsx(
  here(
    "./01_data/raw_data/Hibernation_dataset_4_3_arousal.xlsx"
  ),
  colNames = T,
  sheet = 3
)

paged_table(data, options = list(rows.print = 20))
```

As a summary of the *initial dataset*:

- Number of studies: `r length(unique(data$Study_ID))`
- Number of species: `r length(unique(data$Species))`

For this initial meta-analysis, I am going to include all immune (IM) markers and calculate the standarised mean difference (g; with heteroscedastic population variances) as the effect size of analysis comparing **arousal** *versus* **euthermia**. Positive SMD values indicate higher OS values in *arousal* while negative SMD indicate lower OS values in arousal. This is a summary of the effect size calculation based on the dataset above:

```{r echo = TRUE, eval = TRUE, include = TRUE}
# ES data
data <- readRDS(
  here(
    "./01_data/processed_RDS_data_files/06_GC_metaanalysis_arousal_full_data.RDS"
  )
)
data$obsID <- 1:nrow(data) # observation ID


data$scientific_name <- gsub(
  x = data$Species,
  pattern = "_",
  replacement = " "
)
table(data$scientific_name)

## data for SMD
df_SMD <- data %>%
  filter(!is.na(SMDH)) %>%
  filter(!is.na(SMDH.sv))
```

Summary of SMD effect size and number of observations included in the meta-analysis:

```{r echo = TRUE, eval = TRUE, include = TRUE}
summary(df_SMD$SMDH)
```

- Number of observation with non-missing SMD: `r nrow(df_SMD)`
- Number of studies in the dataset for analysis: `r length(unique(df_SMD$Study_ID))`
- Number of species in the dataset for analysis: `r length(unique(df_SMD$Species))`

## Initial meta-analysis:

Only three observations left. No meta-analysis.