---
title: "Meta-analysis physiological costs of hibernation: Report 3 - GC Hibernation"
author: "Pablo Capilla-Lasheras"
date: "2026-05-04"
output:
  html_document:
    df_print: paged
editor_options: 
  chunk_output_type: console
knit: (function(inputFile, encoding) {
      out_dir <- "../05_reports";
      rmarkdown::render(inputFile,
                        encoding=encoding,
                        output_dir=file.path(dirname(inputFile), out_dir))})
---

```{r setup, include = FALSE}
# setup
knitr::opts_chunk$set(
  include = TRUE,
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)

library(here)
library(orchaRd)
library(openxlsx)
library(metafor)
library(tidyverse)
library(ggokabeito)
library(orchaRd)
library(rmarkdown)
library(knitr)
library(kableExtra)
library(gtsummary)
library(ggtree)
```

This [R Markdown](http://rmarkdown.rstudio.com) Notebook contains the initial meta-analysis for the project 'physiological 'costs' of hibernation. Specifically, the meta-analysis of hibernation

## Dataset and calculation of effect sizes

The dataset for this meta-analysis has been collected from the literature and curated by Pablo Burraco, Marko Prokic and Tamara Petrovic. It contains data on physiological effects of hibernation in different systems. Here a quick snapshot of the first lines of the dataset:

```{r echo = TRUE, eval = TRUE, include = TRUE}
# dataset with raw data
data <- openxlsx::read.xlsx(
  here(
    "./01_data/raw_data/Burraco_et_al_hibernation_hibernation.xlsx"
  ),
  colNames = T,
  sheet = 3
)

paged_table(data, options = list(rows.print = 20))
```

As a summary of the *initial dataset*:

- Number of studies: `r length(unique(data$StudyID))`
- Number of species: `r length(unique(data$Species))`

For this initial meta-analysis, I am going to include all glucocorticoid (GC) markers and calculate the standarised mean difference (g; with heteroscedastic population variances) as the effect size of analysis comparing **hibernation** *versus* **euthermia**. Positive SMD values indicate higher GC values in *hibernation* while negative SMD indicate lower GC values in hibernation. This is a summary of the effect size calculation based on the dataset above:

```{r echo = TRUE, eval = TRUE, include = TRUE}
# ES data
data <- readRDS(
  here(
    "./01_data/processed_RDS_data_files/03_GC_metaanalysis_hibernation_full_data.RDS"
  )
)
data$obsID <- 1:nrow(data) # observation ID

data$scientific_name <- gsub(
  x = data$Species,
  pattern = "_",
  replacement = " "
)

data$scientific_name_phylo <- data$scientific_name

## data for SMD
df_SMD <- data %>%
  filter(!is.na(SMDH)) %>%
  filter(!is.na(SMDH.sv))

# matrix with phylogentic correlations
phylo_cor <- readRDS(
  here(
    "./01_data/processed_RDS_data_files/03_GC_phylogenetic_correlations_hibernation.RDS"
  )
)

phylo_tree <- readRDS(
  here(
    "./01_data/processed_RDS_data_files/03_GC_phylogenetic_tree_hibernation.RDS"
  )
)

tree <- ggtree(phylo_tree, size = 1.2) +
  xlim_tree(c(0, 35)) +
  geom_tiplab(
    aes(
      label = paste0(
        "italic('",
        sub(x = label, pattern = "_", replacement = " "),
        "')"
      )
    ),
    parse = TRUE,
    offset = 0.5,
    size = 5,
    hjust = 0
  )

ggsave(
  filename = here(
    "./03_plots/01_hibernation/00c_GC_phylo_tree.png"
  ),
  plot = tree,
  device = "png",
  height = 150,
  width = 90,
  units = "mm"
)

ggsave(
  filename = here(
    "./03_plots/01_hibernation/00c_GC_phylo_tree.pdf"
  ),
  plot = tree,
  height = 150,
  width = 90,
  units = "mm"
)
```

Summary of SMD effect size and number of observations included in the meta-analysis:

```{r echo = TRUE, eval = TRUE, include = TRUE}
summary(df_SMD$SMDH)
print(tree)
```

- Number of observation with non-missing SMD: `r nrow(df_SMD)`
- Number of studies in the dataset for analysis: `r length(unique(df_SMD$StudyID))`
- Number of species in the dataset for analysis: `r length(unique(df_SMD$Species))`

## Initial meta-analysis:

This initial model contains only the intercept and four random effects to account for phylogenetic variation and variation between and within studies beyond sampling variance.

```{r echo = TRUE, eval = TRUE, include = TRUE}
## calculate var-cover matrix to control for correlated effect sizes within study and tissue
V <- vcalc(
  vi = SMDH.sv,
  cluster = StudyID,
  subgroup = Tissue,
  grp1 = Hib_M,
  grp2 = Euthermia_M,
  w1 = Hib_N,
  w2 = Euthermia_N,
  data = df_SMD
)

# model
model_SMD <- rma.mv(
  yi = SMDH,
  V = V,
  mods = ~1,
  random = list(
    ~ 1 | scientific_name_phylo,
    ~ 1 | Species,
    ~ 1 | obsID,
    ~ 1 | StudyID
  ),
  data = df_SMD,
  R = list(scientific_name_phylo = phylo_cor),
  method = "REML",
  control = list(optimizer = "BFGS"),
  verbose = F
)
```

```{r echo = TRUE, eval = TRUE, include = TRUE}
saveRDS(
  object = model_SMD,
  file = here(
    './04_models/01_hibernation/01c_metamodel_hibernation_GC.RDS'
  )
)
model_SMD <- readRDS(
  file = here(
    './04_models/01_hibernation/01c_metamodel_hibernation_GC.RDS'
  )
)
summary(model_SMD)
i2_SMD <- round(i2_ml(model_SMD), digits = 4) # heterogeneity random effects
```

Proportion of variation left after accounting for sampling variance and proportion accounted by each random effect:

```{r echo = TRUE, eval = TRUE, include = TRUE}
knitr::kable(
  x = data.frame(
    Variable = names(i2_SMD),
    i2_value_percentage = as.numeric(round(i2_SMD, digits = 3))
  )
) %>%
  kableExtra::kable_styling(full_width = F)
```

### Plot of results of intercept-only model:

```{r echo = TRUE, eval = TRUE, include = TRUE}
overall_plot <- orchard_plot(
  object = model_SMD,
  mod = "1",
  group = 'StudyID',
  k.pos = 5,
  trunk.size = 1,
  alpha = 0.2,
  cb = FALSE,
  xlab = "Standardised Mean Difference",
  transfm = "none"
) +
  scale_fill_okabe_ito() +
  scale_color_okabe_ito() +
  theme(
    axis.text = element_text(size = 12),
    axis.text.y = element_text(size = 12, angle = 0),
    axis.title = element_text(size = 16)
  ) +
  scale_y_continuous(
    limits = c(-10, 10),
    breaks = seq(-10, 10, 5),
    label = seq(-10, 10, 5)
  )

overall_plot

ggsave(
  filename = here(
    "./03_plots/01_hibernation/01c_overall_plot_hibernation_GC.png"
  ),
  plot = overall_plot,
  device = "png",
  height = 110,
  width = 185,
  units = "mm"
)

ggsave(
  filename = here(
    "./03_plots/01_hibernation/01c_overall_plot_hibernation_GC.pdf"
  ),
  plot = overall_plot,
  height = 110,
  width = 185,
  units = "mm"
)
```

### Publication bias: small-study effects

```{r echo = TRUE, eval = TRUE, include = TRUE}
df_SMD$inv_eff_ss <- (1 / df_SMD$Hib_N) + (1 / df_SMD$Euthermia_N)

model_SMD_pb <- rma.mv(
  yi = SMDH,
  V = V,
  mods = ~ 1 + sqrt(inv_eff_ss),
  random = list(
    ~ 1 | scientific_name_phylo,
    ~ 1 | Species,
    ~ 1 | obsID,
    ~ 1 | StudyID
  ),
  data = df_SMD,
  R = list(scientific_name_phylo = phylo_cor),
  method = "REML",
  control = list(optimizer = "BFGS"),
  verbose = F
)
```

```{r echo = TRUE, eval = TRUE, include = TRUE}
saveRDS(
  object = model_SMD_pb,
  file = here(
    './04_models/01_hibernation/06c_pb_metamodel_hibernation_GC.RDS'
  )
)
model_SMD_pb <- readRDS(
  file = here(
    './04_models/01_hibernation/06c_pb_metamodel_hibernation_GC.RDS'
  )
)

summary(model_SMD_pb)
```

### Publication bias: time-lag effects

```{r echo = TRUE, eval = TRUE, include = TRUE}
df_SMD$year_scl <- scale(df_SMD$Year)

model_SMD_tl <- rma.mv(
  yi = SMDH,
  V = V,
  mods = ~ 1 + year_scl,
  random = list(
    ~ 1 | scientific_name_phylo,
    ~ 1 | Species,
    ~ 1 | obsID,
    ~ 1 | StudyID
  ),
  data = df_SMD,
  R = list(scientific_name_phylo = phylo_cor),
  method = "REML",
  control = list(optimizer = "BFGS"),
  verbose = F
)
```

```{r echo = TRUE, eval = TRUE, include = TRUE}
saveRDS(
  object = model_SMD_tl,
  file = here(
    './04_models/01_hibernation/07c_timelag_metamodel_hibernation_GC.RDS'
  )
)
model_SMD_tl <- readRDS(
  file = here(
    './04_models/01_hibernation/07c_timelag_metamodel_hibernation_GC.RDS'
  )
)

summary(model_SMD_tl)
```

```{r echo = TRUE, eval = TRUE, include = TRUE}
results <- mod_results(model_SMD_tl, mod = "year_scl", group = "StudyID")
orchaRd::bubble_plot(
  model_SMD_tl,
  mod = 'year_scl',
  group = 'StudyID',
  alpha = 0.2,
  cb = FALSE,
  xlab = "Year of publication (scaled)",
  transfm = "none",
) +
  theme(
    axis.text = element_text(size = 12),
    axis.text.y = element_text(size = 12, angle = 0),
    axis.title = element_text(size = 16)
  )
```


## Meta-analysis including thermoregulation type as a moderator:

Number of observations for each type:

```{r echo = TRUE, eval = TRUE, include = TRUE}
table(df_SMD$Thermoregulation)
```

```{r echo = TRUE, eval = TRUE, include = TRUE}
model_SMD_thermoregulation <- rma.mv(
  yi = SMDH,
  V = V,
  mods = ~ Thermoregulation - 1,
  random = list(
    ~ 1 | scientific_name_phylo,
    ~ 1 | Species,
    ~ 1 | obsID,
    ~ 1 | StudyID
  ),
  data = df_SMD,
  R = list(scientific_name_phylo = phylo_cor),
  method = "REML",
  control = list(optimizer = "BFGS"),
  verbose = F
)
```

```{r echo = TRUE, eval = TRUE, include = TRUE}
saveRDS(
  object = model_SMD_thermoregulation,
  file = here(
    './04_models/01_hibernation/03c_metareg_thermoregulation_hibernation_GC.RDS'
  )
)
model_SMD_thermoregulation <- readRDS(
  file = here(
    './04_models/01_hibernation/03c_metareg_thermoregulation_hibernation_GC.RDS'
  )
)
summary(model_SMD_thermoregulation)
```

### Plot of results of thermoregulation meta-regression:

```{r echo = TRUE, eval = TRUE, include = TRUE}
thermo_plot <- orchard_plot(
  object = model_SMD_thermoregulation,
  mod = 'Thermoregulation',
  group = 'StudyID',
  k.pos = 5,
  trunk.size = 1,
  alpha = 0.2,
  cb = FALSE,
  xlab = "Standardised Mean Difference",
  transfm = "none"
) +
  scale_fill_okabe_ito() +
  scale_color_okabe_ito() +
  theme(
    axis.text = element_text(size = 12),
    axis.text.y = element_text(size = 12, angle = 0),
    axis.title = element_text(size = 16)
  ) +
  scale_y_continuous(
    limits = c(-10, 10),
    breaks = seq(-10, 10, 5),
    label = seq(-10, 10, 5)
  )

thermo_plot

ggsave(
  filename = here(
    "./03_plots/01_hibernation/03c_thermoregulation_plot_GC.png"
  ),
  plot = thermo_plot,
  device = "png",
  height = 120,
  width = 185,
  units = "mm"
)

ggsave(
  filename = here(
    "./03_plots/01_hibernation/03c_thermoregulation_plot_GC.pdf"
  ),
  plot = thermo_plot,
  height = 120,
  width = 185,
  units = "mm"
)
```

